---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
```

# avesperu
## <a href='https://github.com/PaulESantos/avesperu'><img src='man/figures/avesperu.svg' align="right" height="370" width="180" /></a>
<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/avesperu)](https://CRAN.R-project.org/package=avesperu)
[![R-CMD-check](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml)
[![](http://cranlogs.r-pkg.org/badges/grand-total/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
[![](http://cranlogs.r-pkg.org/badges/last-week/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
<!-- badges: end -->

The `avesperu` package opens the door to Peru's extraordinary birds, encompassing a dataset of 1,901 species, highlighting its status as one of the world's most diverse bird-rich countries. This comprehensive R package now leverages the meticulously updated "List of the Birds of Peru" by M. A. Plenge, which has been revised and expanded to reflect the latest data as of 06 - 03 - 2024. However, it's important to recognize that avian taxonomy is dynamic, undergoing significant changes over time.

The species list adheres to the taxonomic classification endorsed by the South American Checklist Committee (SACC), which assigns species codes reflecting various aspects such as residency status, endemism, migratory behavior, vagrancy, introductions, extirpation, and hypothetical presence. As of now, SACC is in the process of recognizing certain subspecies as full species, a development that will inevitably alter the species count.

Of particular note, the "H" category, representing hypothetical species, now holds the fifth position in South America's ranking, following Argentina, Bolivia, Colombia, and Ecuador. However, this ranking may potentially decrease further with the publication of well-documented records, underscoring the ongoing evolution and refinement of avian taxonomy.

```{r echo=FALSE}
tibble::tibble(year = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024),
               species = c(1491, 1678, 1800, 1800, 
                           1820, 1851, 1870, 1874, 
                           1888, 1892, 1901)) |> 
  ggplot2::ggplot(ggplot2::aes(year,
             species)) +
  ggplot2::geom_line(ggplot2::aes(group = 1)) +
  ggplot2::geom_point(color = "red") +
  ggplot2::labs(x = "",
       y = "Number of species") +
  ggplot2::scale_x_continuous(guide = ggplot2::guide_axis(n.dodge = 2),
  breaks = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024),
  labels = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024)) +
  ggplot2::theme_bw() +
  ggplot2::coord_flip()
```


Suggested citation:

```{r}
citation("avesperu")
```

## Installation

You can install the `avesperu` package from CRAN using:

```r
install.packages("avesperu")
# or
pak::pak("avesperu")
```

Also you can install the development version of `avesperu` like so:

``` r
pak::pak("PaulESantos/avesperu")
```

## Usage

Here's a quick example of how to use the `avesperu` package:



```{r}
library(avesperu)

splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Thraupisa palamarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.05)

```

- The package not only provides access to the list of bird species recorded in Peru but also excels in resolving potential typos or variations in species names through fuzzy matching. It ensures accurate retrieval by intelligently recognizing and accommodating slight discrepancies in the input names, making it a robust tool for working with diverse and sometimes inconsistent datasets.

```{r}
splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Thraupisa palamarum",
            "Thraupisa palamarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.1)

```
