---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# avesperu <a href='https://github.com/PaulESantos/avesperu'><img src='man/figures/avesperu.svg' align="right" height="200" width="200" /></a>


<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/avesperu)](https://CRAN.R-project.org/package=avesperu)
[![R-CMD-check](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml)
[![](http://cranlogs.r-pkg.org/badges/grand-total/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
[![](http://cranlogs.r-pkg.org/badges/last-week/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->

The `avesperu` package provides access to the most up-to-date and comprehensive dataset on Peru's avian diversity. As of **April 5, 2025**, the list includes **1,912** bird species, reflecting significant taxonomic changes and updated validations based on recent scientific publications, photographs, and sound recordings deposited in accredited institutions. The classification follows the guidelines set by the South American Checklist Committee (SACC).

### Species Categories

Each species in the dataset is classified into one of the following categories, reflecting its status in Peru:

- X Resident: 1,542 species
- E Endemic: 119 species
- NB Migratory (non-breeding): 138 species
- V Vagrant: 84 species
- IN Introduced: 3 species
- EX Extirpated: 0 species
- H Hypothetical: 26 species

This results in a total of 1,912 species, showcasing Peru's extraordinary bird diversity and the ongoing refinement of its avifaunal checklist.

## Features

The `avesperu` package is designed to streamline access to this data for researchers, conservationists, and bird enthusiasts alike. It provides:

- A comprehensive and updated bird species dataset following the latest SACC classification.

- Taxonomy validation tools, ensuring consistency with international standards.

- Fuzzy matching capabilities for improved species name retrieval and validation.


```{r echo=FALSE}
tibble::tibble(year = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025),
               species = c(1491, 1678, 1800, 1800, 
                           1820, 1851, 1870, 1874, 
                           1888, 1892, 1909, 1912)) |> 
  ggplot2::ggplot(ggplot2::aes(year,
             species)) +
  ggplot2::geom_line(ggplot2::aes(group = 1)) +
  ggplot2::geom_point(color = "red") +
  ggplot2::labs(x = "",
       y = "Number of species") +
  ggplot2::scale_x_continuous(guide = ggplot2::guide_axis(n.dodge = 2),
  breaks = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025),
  labels = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025)) +
  ggplot2::theme_bw() +
  ggplot2::coord_flip()
```


### Insights and Trends

The chart shows the steady increase in the number of bird species recorded in Peru from 1968 to 2025, reflecting continuous research and improvements in taxonomic resolution:

- A substantial jump occurred between 1968 and 1980, with 187 new species recorded.

- In recent years, updates have slowed but continued to increase steadily, reflecting meticulous reviews of published records and taxonomic refinements.

### Categorical Summary

- Most species are resident (X), accounting for 1,542 species.

- Endemic species (E) have increased to 119, reinforcing Peru’s status as a center of avian endemism.

- The hypothetical category (H) includes 26 species, pending stronger evidence for full inclusion.

## Suggested citation:

```{r}
citation("avesperu")
```

## Installation

You can install the `avesperu` package from CRAN using:

```r
install.packages("avesperu")
# or
pak::pak("avesperu")
```

Also you can install the development version of `avesperu` like so:

``` r
pak::pak("PaulESantos/avesperu")
```

## Usage

Here's a quick example of how to use the `avesperu` package:



```{r}
library(avesperu)

splist <- c("Falco sparverius",
            "Tinamus osgoodi",
            "Crypturellus soui",
            "Thraupis palmarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist)

```

- The package not only provides access to the list of bird species recorded in Peru but also excels in resolving potential typos or variations in species names through fuzzy matching. It ensures accurate retrieval by intelligently recognizing and accommodating slight discrepancies in the input names, making it a robust tool for working with diverse and sometimes inconsistent datasets.

```{r}
splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Tinamus guttatus",
            "Tinamus guttattus",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.05)

```
