---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# avesperu

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/avesperu)](https://CRAN.R-project.org/package=avesperu)
[![R-CMD-check](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The `avesperu` package provides access to the list of bird species recorded in Peru, as documented by M. A. Plenge in the "List of the Birds of Peru." This package allows you to retrieve information about bird species in Peru, including their scientific names, English and Spanish common names, taxonomic order and family, and status.

## Installation

You can install the development version of avesperu like so:

``` r
pak::pak("PaulESantos/avesperu")
```

## Usage

Here's a quick example of how to use the `avesperu` package:



```{r}
library(avesperu)


splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Thraupisa palamarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.05)

```

