---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# avesperu <a href='https://github.com/PaulESantos/avesperu'><img src='man/figures/avesperu.svg' align="right" height="250" width="220" /></a>


<!-- badges: start -->
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-green.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![CRAN status](https://www.r-pkg.org/badges/version/avesperu)](https://CRAN.R-project.org/package=avesperu)
[![R-CMD-check](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PaulESantos/avesperu/actions/workflows/R-CMD-check.yaml)
[![](http://cranlogs.r-pkg.org/badges/grand-total/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
[![](http://cranlogs.r-pkg.org/badges/last-week/avesperu?color=green)](https://cran.r-project.org/package=avesperu)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
<!-- badges: end -->


The `avesperu` package provides access to the most up-to-date and comprehensive dataset on Peru's avian diversity. As of **February 6, 2025**, the list includes **1,911 bird species**, reflecting significant advancements in taxonomy and the validation of species records. These updates are based on articles, photographs, and sound recordings archived in accredited institutions, and follow the classification endorsed by the South American Checklist Committee (SACC).

### Species Categories

Each species in the dataset is classified into one of the following categories, reflecting its status in Peru:

- **X Resident**: 1,542 species
- **E Endemic**: 117 species
- **NB Migratory**: 138 species
- **V Vagrant**: 83 species
- **IN Introduced**: 3 species
- **EX Extirpated**: 0 species
- **H Hypothetical**: 26 species

This results in a total of 1,911 species, demonstrating Peru's incredible avian richness.

### Features

- The `avesperu` package is designed to streamline access to this data for researchers, conservationists, and bird enthusiasts alike. It provides:

- A comprehensive and updated bird species dataset following the latest SACC classification.

- Taxonomy validation tools, ensuring consistency with international standards.

- Fuzzy matching capabilities for improved species name retrieval and validation.

```{r echo=FALSE}
tibble::tibble(year = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025),
               species = c(1491, 1678, 1800, 1800, 
                           1820, 1851, 1870, 1874, 
                           1888, 1892, 1909, 1911)) |> 
  ggplot2::ggplot(ggplot2::aes(year,
             species)) +
  ggplot2::geom_line(ggplot2::aes(group = 1)) +
  ggplot2::geom_point(color = "red") +
  ggplot2::labs(x = "",
       y = "Number of species") +
  ggplot2::scale_x_continuous(guide = ggplot2::guide_axis(n.dodge = 2),
  breaks = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025),
  labels = c(1968, 1980, 2001, 2007, 
                        2010, 2015, 2020, 2021,
                        2022, 2023, 2024, 2025)) +
  ggplot2::theme_bw() +
  ggplot2::coord_flip()
```


Here’s an enhanced analysis of the provided information, focusing on the trends and implications for Peru's avian biodiversity:

- The chart depicts the steady increase in the number of bird species recorded in Peru from 1968 to 2025, showcasing the country’s remarkable progress in avian taxonomy and biodiversity documentation. Over the decades, several key trends emerge:

### Significant Milestones:

The species count grew substantially between 1968 (1,491 species) and 1980 (1,678 species), reflecting early efforts in exploration and classification.
Post-2000, the growth rate appears more stable, with key updates in 2010, 2020, and culminating in 2025 with 1,911 species.


### Categorical Analysis:

- Most of the species belong to the "Resident" (X) category, accounting for **1,542 species**, while **118 species** are classified as  **endemic,** showcasing Peru's unique biodiversity.

- Hypothetical species ("H") contribute 26 species, emphasizing the importance of continued validation efforts to solidify their inclusion in the list.



## Suggested citation:

```{r}
citation("avesperu")
```

## Installation

You can install the `avesperu` package from CRAN using:

```r
install.packages("avesperu")
# or
pak::pak("avesperu")
```

Also you can install the development version of `avesperu` like so:

``` r
pak::pak("PaulESantos/avesperu")
```

## Usage

Here's a quick example of how to use the `avesperu` package:



```{r}
library(avesperu)

splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Thraupisa palamarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.05)

```

- The package not only provides access to the list of bird species recorded in Peru but also excels in resolving potential typos or variations in species names through fuzzy matching. It ensures accurate retrieval by intelligently recognizing and accommodating slight discrepancies in the input names, making it a robust tool for working with diverse and sometimes inconsistent datasets.

```{r}
splist <- c("Falco sparverius",
            "Tinamus osgodi",
            "Crypturellus sooui",
            "Thraupisa palamarum",
            "Thraupisa palamarum",
            "Thamnophilus praecox")

search_avesperu(splist = splist, max_distance = 0.05)

```
