---
title: "Comparison between aves_peru_2025_v5 and aves_peru_2026_v1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison between aves_peru_2025_v5 and aves_peru_2026_v1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This vignette highlights the differences and updates between the previous dataset `aves_peru_2025_v5` and the newly released dataset `aves_peru_2026_v1` in the `avesperu` package.

```{r setup}
library(avesperu)
```

## Summary Comparison

Let's compare the total number of species and the distribution of their statuses.

```{r comparison}
# 2025 version
table(aves_peru_2025_v5$status)

# 2026 version
table(aves_peru_2026_v1$status)
```

## Additions and Removals

Identify the species that were added or removed in the 2026 update.

```{r additions-removals}
added_species <- setdiff(aves_peru_2026_v1$scientific_name, aves_peru_2025_v5$scientific_name)
removed_species <- setdiff(aves_peru_2025_v5$scientific_name, aves_peru_2026_v1$scientific_name)

length(added_species)
length(removed_species)

# Display some newly added species
head(added_species)
```

## Taxonomic Changes

Check if any families or orders have updated species counts or additions.
```{r taxonomy}
# 2025 Families count
length(unique(aves_peru_2025_v5$family_name))

# 2026 Families count
length(unique(aves_peru_2026_v1$family_name))
```

This ensures users can easily transition their workflows and verify taxonomic consistency between these iterations.
